The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2016-04-08 script/en/chapters/bitciphers.tex
CT Book ed 12: Update of the bitcipher chapter, and many small typos like English '' instead of \glqq in German quotes.

%BERM0 Changed when bc becoming part of the CT book.
%BERM1 Transformed to comment when bc becoming part of the CT book: There it already was in script-en.tex.
%BERM2 Transformed to comment when bc becoming part of the CT book: Added there to script-en.tex.
%BERM3 Redefined new commands already existing in moderncryptography.tex when bc becoming part of the CT book
%BERM3 (see http://tex.stackexchange.com/questions/36175/what-do-newcommand-renewcommand-and-providecommand-do-and-how-do-they-differ).

%BERM1 \documentclass[11pt,a4paper,leqno]{report}
%BERM1 \usepackage[latin1]{inputenc}
%BERM2 \usepackage[T1]{fontenc}
%BERM2 \usepackage{amsmath,amssymb,amscd,latexsym}
%BERM2 \usepackage{color}
%BERM1 \usepackage{hyperref}
%BERM1 \usepackage{makeidx}
%BERM1 \usepackage[pdftex]{graphicx}
%BERM1 \usepackage{float}

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
% Pommerenings Specialities
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% \newcommand*{\C}{\mathbb{C}}   %BERM3
\renewcommand*{\C}{\mathbb{C}}   %BERM3
\newcommand*{\F}{\mathbb{F}}
\newcommand*{\M}{\mathbb{M}}
\renewcommand*{\N}{\mathbb{N}}   %BERM3
\renewcommand*{\Q}{\mathbb{Q}}   %BERM3
\renewcommand*{\R}{\mathbb{R}}   %BERM3
\renewcommand*{\Z}{\mathbb{Z}}   %BERM3
\newcommand*{\lsb}{\operatorname{lsb}}
\newcommand*{\Oh}{\operatorname{O}}

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
% CrypTool-Spezialitäten
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

%BERM1 \newtheorem{definition}{Definition}[section]
%BERM1 \newtheorem{theorem}{Theorem}[section]
%BERM1 \newenvironment{proof}[1]{\noindent\textbf{Proof #1} \\}{\hfill$\Box$\par}
%BERM1 \newenvironment{example}[1]{\noindent\textbf{Example#1}}{}
%BERM1 \newenvironment{remark}[1]{\noindent\textbf{Remark#1}}{}
%BERM1 \setcounter{secnumdepth}{4}
%BERM1 \floatstyle{ruled}
%BERM1 \newfloat{sagecode}{!ht}{loc}[chapter]
%BERM1 \floatname{sagecode}{SageMath sample}

\sloppy
\frenchspacing
%BERM1 \makeindex
%BERM1 \begin{document}

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
% B i t b l o c k -  a n d  B i t s t r e a m - E n c r y p t i o n
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

%BERM0 \setcounter{chapter}{98}  % ====> Dummy - beim Einfügen ins CT-Skript löschen <=====
\setcounter{theorem}{0}
\setcounter{definition}{0}

\newpage %BERM0 
\hypertarget{Chapter_BitCiphers}{}   %BERM0
\chapter{Introduction to Bitblock and Bitstream Ciphers}
\label{Chapter_BitCiphers}
(Klaus Pommerening, 2015) \\

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage

\section{Boolean Functions}\label{s-bool-fct}

\subsection{Bits and Their Composition}\label{s-bool-bit}

@@

For a comprehensive treatment see \cite{CuSt} or the two articles
by Claude Carlet in \cite{CrHa}.

@@

\begin{table}[hbpt]
\begin{center}
  \begin{tabular}{|cccc|l|ll|l|} \hline 
    $a$&$b$&$c$&$d$& ANF        & \multicolumn{2}{l|}{logical operation}  & CNF               \\ \hline\hline
     0 & 0 & 0 & 0 & $0$        & False     & constant                    & $x \wedge \neg x$ \\ \hline
     1 & 0 & 0 & 0 & $1$        & True      & constant                    & $x \vee \neg x$   \\ \hline
     0 & 1 & 0 & 0 & $x$        & $x$       & projection                  & $x$               \\ \hline
     1 & 1 & 0 & 0 & $1+x$      & $\neg x$  & negation                    & $\neg x$          \\ \hline\hline
     0 & 0 & 1 & 0 & $y$        & $y$       & projection                  & $y$               \\ \hline
     1 & 0 & 1 & 0 & $1+y$      & $\neg y$  & negation                    & $\neg y$          \\ \hline
     0 & 1 & 1 & 0 & $x+y$      & $x$ XOR $y$ & XOR & $(x \vee y) \wedge (\neg x \vee \neg y)$ \\ \hline
     1 & 1 & 1 & 0 & $1+x+y$    &$x\Longleftrightarrow y$& equivalence &$(\neg x\vee y)\wedge(x\vee\neg y)$\\ \hline\hline
     0 & 0 & 0 & 1 & $xy$       & $x\wedge y$ & AND                       & $x\wedge y$       \\ \hline
     1 & 0 & 0 & 1 & $1+xy$     & $\neg(x\wedge y)$ & NAND                & $(\neg x)\vee(\neg y)$\\ \hline
     0 & 1 & 0 & 1 & $x+xy$     & $x\wedge(\neg y)$ &                     & $x\wedge(\neg y)$ \\ \hline
     1 & 1 & 0 & 1 & $1+x+xy$   & $x\Longrightarrow y$& implication       & $(\neg x)\vee y$  \\ \hline\hline
     0 & 0 & 1 & 1 & $y+xy$     & $(\neg x)\wedge y$ &                    & $(\neg x)\wedge y$\\ \hline
     1 & 0 & 1 & 1 & $1+y+xy$   & $x\Longleftarrow y$&  implication       & $x \vee (\neg y)$ \\ \hline
     0 & 1 & 1 & 1 & $x+y+xy$   & $x \vee y$ & OR                         & $x \vee y$        \\ \hline
     1 & 1 & 1 & 1 & $1+x+y+xy$ & $\neg(x\vee y)$& NOR              & $(\neg x)\wedge(\neg y)$\\ \hline
  \end{tabular}
\end{center}
\caption{The 16 operations on two bits (= Boolean functions of 2 variables),
   using Table~\ref{t-bool-trf}}\label{tab-bool-2}
\end{table}

@@

For a concrete example with $n = l = 4$ we consider the Boolean
map\footnote{%
   By the way $f$ is the S-box\index{S-box} $\mathrm{S}_0$
   of {\sc Lucifer}\index{Lucifer}, a precursor of DES\index{DES}
   developed around 1970.
}
$f$ that is given by the values in Table~\ref{tab-bool-A1}, and
define the bitblock cipher
\[
    F\!: \F_2^4 \times \F_2^4 \longrightarrow \F_2^4 \quad
    \text{by } F(a,k) := f(a + k).
\]
SageMath sample~\ref{Sage-code-Luc-S0} defines this Boolean map
$f =$ {\tt S0}, using the classes {\tt BoolF} and {\tt BoolMap}
from Appendix~\ref{ss-bool-class}. The columns of the (implicit)
defining matrix contain just the values of the map, as they are
also given in the column $y = f(x)$ of Table~\ref{tab-bool-A1}.
A sample evaluation illustrates this (for the third column,
representing the argument {\tt 0010}).

@@

\begin{table}
\begin{center}
\begin{tabular}{|cccc|c|} \hline
  $a$  & $b$  & $b'$ & $c$  & $\alpha(a)+\beta(c)$ \\ \hline
  0000 & 1000 & 0010 & 0011 & 1 \\
  0001 & 1001 & 0110 & 0111 & 1 \\
  0010 & 1010 & 0011 & 0010 & 0 \\
  0011 & 1011 & 0001 & 0000 & 1 \\
  0100 & 1100 & 1001 & 1000 & 1 \\
  0101 & 1101 & 0100 & 0101 & 1 \\
  0110 & 1110 & 0101 & 0100 & 1 \\
  0111 & 1111 & 1000 & 1001 & 1 \\
  1000 & 0000 & 1100 & 1101 & 1 \\
  1001 & 0001 & 1111 & 1110 & 1 \\
  1010 & 0010 & 0111 & 0110 & 1 \\
  1011 & 0011 & 1010 & 1011 & 1 \\
  1100 & 0100 & 1110 & 1111 & 1 \\
  1101 & 0101 & 1101 & 1100 & 1 \\
  1110 & 0110 & 1011 & 1010 & 1 \\
  1111 & 0111 & 0000 & 0001 & 0 \\
  \hline
\end{tabular}
\end{center}
\caption{A linear relation for the key bits}\label{tab-bool-linrel}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|cccccccccccccccc|} \hline
     & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 &10 &11 &12 &13 &14 &15 \\ \hline
   0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
   1 & 0 &$-\frac{1}{4}$&$-\frac{1}{4}$& 0 & 0 &$-\frac{1}{4}$&$-\frac{1}{4}$& 0 & 0 &$-\frac{1}{4}$&$-\frac{1}{4}$& 0 & 0 &$\frac{3}{4}$&$-\frac{1}{4}$& 0 \\
   2 & 0 &$\frac{1}{4}$& 0 & $-\frac{1}{4}$ &$-\frac{1}{2}$& $-\frac{1}{4}$ & 0 & $-\frac{1}{4}$ & $-\frac{1}{4}$ &$\frac{1}{2}$& $-\frac{1}{4}$ & 0 &$\frac{1}{4}$& 0 & $-\frac{1}{4}$ & 0 \\
   3 & 0 &$\frac{1}{2}$&$\frac{1}{4}$& $-\frac{1}{4}$ &$\frac{1}{2}$& 0 &$\frac{1}{4}$& $-\frac{1}{4}$ & $-\frac{1}{4}$ & $-\frac{1}{4}$ & 0 & 0 &$\frac{1}{4}$&$\frac{1}{4}$& 0 & 0 \\
   4 & 0 & 0 &$-\frac{1}{2}$& 0 & 0 & 0 & 0 &$-\frac{1}{2}$&$\frac{1}{4}$& $-\frac{1}{4}$ & $-\frac{1}{4}$ & $-\frac{1}{4}$ &$\frac{1}{4}$& $-\frac{1}{4}$ &$\frac{1}{4}$&$\frac{1}{4}$\\
   5 & 0 &$\frac{1}{4}$&$\frac{1}{4}$&$\frac{1}{2}$& 0 &$\frac{1}{4}$& $-\frac{1}{4}$ & 0 &$\frac{1}{4}$& 0 &$-\frac{1}{2}$&$\frac{1}{4}$&$\frac{1}{4}$& 0 & 0 & $-\frac{1}{4}$ \\
   6 & 0 &$\frac{1}{4}$& 0 &$\frac{1}{4}$& 0 &$\frac{1}{4}$& 0 &$\frac{1}{4}$& 0 &$\frac{1}{4}$& 0 & 2 & 0 &$\frac{1}{4}$& 0 &$\frac{1}{4}$\\
   7 & 0 & 0 &$\frac{1}{4}$& $-\frac{1}{4}$ & 0 & 0 & 2 & $-\frac{1}{4}$ & 0 & 0 &$\frac{1}{4}$& $-\frac{1}{4}$ & 0 & 0 &$\frac{1}{4}$& $-\frac{1}{4}$ \\
   8 & 0 & 0 & $-\frac{1}{4}$ &$\frac{1}{4}$& $-\frac{1}{4}$ &$\frac{1}{4}$& 0 & 0 &$-\frac{1}{2}$& 0 &$\frac{1}{4}$&$\frac{1}{4}$&$\frac{1}{4}$&$\frac{1}{4}$&$\frac{1}{2}$& 0 \\
   9 & 0 &$\frac{1}{4}$& 0 &$\frac{1}{4}$& $-\frac{1}{4}$ &$-\frac{1}{2}$&$\frac{1}{4}$& 0 & 0 & $-\frac{1}{4}$ & 0 & $-\frac{1}{4}$ & $-\frac{1}{4}$ & 0 &$\frac{1}{4}$&$-\frac{1}{2}$\\
  10 & 0 & $-\frac{1}{4}$ &$\frac{1}{4}$& 0 & $-\frac{1}{4}$ & 0 & 0 &$\frac{1}{4}$& $-\frac{1}{4}$ &$-\frac{1}{2}$& 0 & $-\frac{1}{4}$ &$\frac{1}{2}$& $-\frac{1}{4}$ & $-\frac{1}{4}$ & 0 \\
  11 & 0 &$\frac{1}{2}$& 0 & 0 & $-\frac{1}{4}$ & $-\frac{1}{4}$ & $-\frac{1}{4}$ &$\frac{1}{4}$&$\frac{1}{4}$& $-\frac{1}{4}$ &$\frac{1}{4}$&$\frac{1}{4}$& 0 & 0 & 0 &$\frac{1}{2}$\\
  12 & 0 & 0 &$\frac{1}{4}$&$\frac{1}{4}$& $-\frac{1}{4}$ &$\frac{1}{4}$& 0 &$-\frac{1}{2}$& $-\frac{1}{4}$ & $-\frac{1}{4}$ & 0 & 0 &$-\frac{1}{2}$& 0 & $-\frac{1}{4}$ &$\frac{1}{4}$\\
  13 & 0 & $-\frac{1}{4}$ &$\frac{1}{2}$& $-\frac{1}{4}$ & $-\frac{1}{4}$ & 0 &$\frac{1}{4}$& 0 &$\frac{1}{4}$& 0 & $-\frac{1}{4}$ & 0 & 0 &$\frac{1}{4}$&$\frac{1}{2}$& 0 \\
  14 & 0 & $-\frac{1}{4}$ &$\frac{1}{4}$&$\frac{1}{2}$&$\frac{1}{4}$&$-\frac{1}{2}$& 0 & $-\frac{1}{4}$ & 0 &$\frac{1}{4}$&$\frac{1}{4}$& 0 &$\frac{1}{4}$& 0 & 0 &$\frac{1}{4}$\\
  15 & 0 & 0 & 0 & 0 &$\frac{1}{4}$& $-\frac{1}{4}$ & $-\frac{1}{4}$ &$\frac{1}{4}$&$-\frac{1}{2}$& 0 &$-\frac{1}{2}$& 0 & $-\frac{1}{4}$ & $-\frac{1}{4}$ &$\frac{1}{4}$&$\frac{1}{4}$\\ \hline
\end{tabular}
\end{center}\caption{Correlation matrix of the S-box $\mathrm{S}_0$ of {\sc Lucifer}\index{Lucifer}.
   Row and column indices are linear forms represented by integers.}\label{tab-bool-corr}
\end{table}

@@

     & 0            &$\frac{1}{16}$& 0            & 0            &$\frac{1}{16}$&$\frac{1}{4}$ &$\frac{1}{16}$\\

@@

Once again we consider a concrete example with $m = 2$ and $q = 4$,
hence $n = 8$. As S-boxes we take the ones from {\sc Lucifer}, $\textrm{S}_0$
at the left, and $\textrm{S}_1$ at the right, see Figure~\ref{fig-bool-par}.
For the left S-box $\textrm{S}_0$ we take the linear relation
with $\alpha \:\hat{=}$ \verb:0001: and $\beta \:\hat{=}$ \verb:1101:,
that we know has probability $p_1 = \frac{7}{8}$, for the right
S-Box $\textrm{S}_1$ we take the relation $(0,0)$ with probability $1$.
The combined linear relation for $f = (\textrm{S}_0,\textrm{S}_1)$ then
also has probability $p = \frac{7}{8}$ and potential $\lambda = \frac{9}{16}$,
and we know that linear cryptanalysis with $N = 5$ pairs of plaintext
and ciphertext has 95\% success probability. We decompose all relevant
bitblocks into bits:
\begin{description}
   \item[plaintext:] $a = (a_0, \ldots, a_7) \in \F_2^8$,
   \item[ciphertext:] $c = (c_0, \ldots, c_7) \in \F_2^8$,
   \item[key:] $k = (k_0, \ldots, k_{15}) \in \F_2^{16}$ where
      $(k_0, \ldots, k_7)$ serves as ``initial key'' (corresponding to
      $k^{(0)}$ in Figure~\ref{fig-bool-par}), and $(k_8, \ldots, k_{15})$
      as ``final key'' (corresponding to $k^{(1)}$).
\end{description}
Then $\alpha(a) = a_3$, $\beta(c) = c_0 + c_1 + c_3$, and
$\kappa(k) = \alpha(k_0, \ldots, k_7) + \beta(k_8, \ldots, k_{15}) = k_3 + k_8 + k_9 + k_{11}$.
Hence the target relation is
\[
     k_3 + k_8 + k_9 + k_{11} = a_3 + c_0 + c_1 + c_3.
\]
We use the key $k =$ \verb:1001011000101110: whose relevant bit is
$k_3 + k_8 + k_9 + k_{11} = 1$, and generate five random pairs of
plaintext and ciphertext, see Table~\ref{tab-bool-BspC}. We see that
for this example Matsui's algorithm guesses the relevant key bit
correctly with no dissentient.

@@

%BERM0: Verschoben an die richtige Stelle.
\begin{figure}[h] % [htbp]
\begin{center}
\begin{picture}(320,50)
  \linethickness{2pt}
  \put(20,20){\line(1,0){260}}
  \put(20,20){\line(0,1){30}}
  \put(20,50){\line(1,0){260}}
  \put(280,20){\line(0,1){30}}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\section{Appendix: Boolean Maps in SageMath}\label{a-bool-sage}

\subsection{What's in SageMath?}

>>>>

% $Id: cryptomethods.tex 3712 2016-03-22 23:36:46Z xesslinger $
% ..............................................................................
% B i t b l o c k -  a n d  B i t s t r e a m - E n c r y p t i o n
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
% Pommerenings Specialities
% ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
% \newcommand*{\C}{\mathbb{C}}   %BERM3
\renewcommand*{\C}{\mathbb{C}}   %BERM3
\newcommand*{\F}{\mathbb{F}}
\newcommand*{\M}{\mathbb{M}}
\renewcommand*{\N}{\mathbb{N}}   %BERM3
\renewcommand*{\Q}{\mathbb{Q}}   %BERM3
\renewcommand*{\R}{\mathbb{R}}   %BERM3
\renewcommand*{\Z}{\mathbb{Z}}   %BERM3
\newcommand*{\lsb}{\operatorname{lsb}}
\newcommand*{\Oh}{\operatorname{O}}

\sloppy
\frenchspacing

\setcounter{theorem}{0}
\setcounter{definition}{0}

\newpage
\hypertarget{Chapter_BitCiphers}{}
\chapter{Introduction to Bitblock and Bitstream Ciphers}
\label{Chapter_BitCiphers}
(Klaus Pommerening, 2015, Updates: Jan 2016, Apr 2016) \\

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\section{Boolean Functions}\label{s-bool-fct}
\subsection{Bits and their Composition}\label{s-bool-bit}

@@

For a comprehensive treatment see \cite{CuSt} or the two articles
by Claude Carlet\footnote{%
  also see his list of publications at
  \href{http://www.math.univ-paris13.fr/~carlet/pubs.html}{\tt http://www.math.univ-paris13.fr/$\sim$carlet/pubs.html}
} in \cite{CrHa}\footnote{%
  Carlet's articles are online at
  \href{http://www.math.univ-paris13.fr/~carlet/chap-fcts-Bool-corr.pdf}{\tt http://www.math.univ-paris13.fr/$\sim$carlet/chap-fcts-Bool-corr.pdf}
  and
  \href{http://www.math.univ-paris13.fr/~carlet/chap-vectorial-fcts-corr.pdf}{\tt http://www.math.univ-paris13.fr/$\sim$carlet/chap-vectorial-fcts-corr.pdf}
}.

@@

\begin{table}[hbpt]
\begin{center}
  \begin{tabular}{|cccc|l|ll|l|} \hline 
    $a$&$b$&$c$&$d$& ANF        & \multicolumn{2}{l|}{logical operation}  & CNF               \\ \hline\hline
     0 & 0 & 0 & 0 & $0$        & False     & constant                    & $x \wedge \neg x$ \\ \hline
     1 & 0 & 0 & 0 & $1$        & True      & constant                    & $x \vee \neg x$   \\ \hline
     0 & 1 & 0 & 0 & $x$        & $x$       & projection                  & $x$               \\ \hline
     1 & 1 & 0 & 0 & $1+x$      & $\neg x$  & negation                    & $\neg x$          \\ \hline\hline
     0 & 0 & 1 & 0 & $y$        & $y$       & projection                  & $y$               \\ \hline
     1 & 0 & 1 & 0 & $1+y$      & $\neg y$  & negation                    & $\neg y$          \\ \hline
     0 & 1 & 1 & 0 & $x+y$      & $x$ XOR $y$ & XOR & $(x \vee y) \wedge (\neg x \vee \neg y)$ \\ \hline
     1 & 1 & 1 & 0 & $1+x+y$    &$x\Longleftrightarrow y$& equivalence &$(\neg x\vee y)\wedge(x\vee\neg y)$\\ \hline\hline
     0 & 0 & 0 & 1 & $xy$       & $x\wedge y$ & AND                       & $x\wedge y$       \\ \hline
     1 & 0 & 0 & 1 & $1+xy$     & $\neg(x\wedge y)$ & NAND                & $(\neg x)\vee(\neg y)$\\ \hline
     0 & 1 & 0 & 1 & $x+xy$     & $x\wedge(\neg y)$ &                     & $x\wedge(\neg y)$ \\ \hline
     1 & 1 & 0 & 1 & $1+x+xy$   & $x\Longrightarrow y$& implication       & $(\neg x)\vee y$  \\ \hline\hline
     0 & 0 & 1 & 1 & $y+xy$     & $(\neg x)\wedge y$ &                    & $(\neg x)\wedge y$\\ \hline
     1 & 0 & 1 & 1 & $1+y+xy$   & $x\Longleftarrow y$&  implication       & $x \vee (\neg y)$ \\ \hline
     0 & 1 & 1 & 1 & $x+y+xy$   & $x \vee y$ & OR                         & $x \vee y$        \\ \hline
     1 & 1 & 1 & 1 & $1+x+y+xy$ & $\neg(x\vee y)$& NOR              & $(\neg x)\wedge(\neg y)$\\ \hline
  \end{tabular}
\end{center}
\caption{The 16 operations on two bits (= Boolean functions of 2 variables),
   using Table~\ref{t-bool-trf} (The order of the first column is
   lexicographic if $a$, $b$, $c$, $d$ are considered in reverse order.)}\label{tab-bool-2}
\end{table}

@@

For a concrete example with $n = l = 4$ we consider the Boolean
map\footnote{%
   By the way $f$ is the S-box\index{S-box} $\mathrm{S}_0$
   of {\sc Lucifer}\index{Lucifer}, a precursor of DES\index{DES}
   developed around 1970.
}
$f$ that is given by the values in Table~\ref{tab-bool-A1}, and
define the bitblock cipher
\[
    F\!: \F_2^4 \times \F_2^4 \longrightarrow \F_2^4 \quad
    \text{by } F(a,k) := f(a + k).
\]
SageMath sample~\ref{Sage-code-Luc-S0} defines this Boolean map
$f =$ {\tt S0}, using the classes {\tt BoolF} and {\tt BoolMap}
from Appendix~\ref{ss-bool-class}.
The columns of the defining matrix (implicit in the SageMath code)
just give the values of the map as they are also found
in the column $y = f(x)$ of Table~\ref{tab-bool-A1}.
(In other words, SageMath sample~\ref{Sage-code-Luc-S0} and
Table~\ref{tab-bool-A1} give equivalent definitions of the map $f$.)
A sample evaluation illustrates this (for the third column,
representing the argument {\tt 0010}).

@@

\begin{table}
\begin{center}
\begin{tabular}{|cccc|c|} \hline
  $a$  & $b$  & $b'$ & $c$  & $\alpha(a)+\beta(c)$ \\ \hline
  0000 & 1000 & 0010 & 0011 & 1 \\
  0001 & 1001 & 0110 & 0111 & 1 \\
  0010 & 1010 & 0011 & 0010 & 0 \\
  0011 & 1011 & 0001 & 0000 & 1 \\
  0100 & 1100 & 1001 & 1000 & 1 \\
  0101 & 1101 & 0100 & 0101 & 1 \\
  0110 & 1110 & 0101 & 0100 & 1 \\
  0111 & 1111 & 1000 & 1001 & 1 \\
  1000 & 0000 & 1100 & 1101 & 1 \\
  1001 & 0001 & 1111 & 1110 & 1 \\
  1010 & 0010 & 0111 & 0110 & 1 \\
  1011 & 0011 & 1010 & 1011 & 1 \\
  1100 & 0100 & 1110 & 1111 & 1 \\
  1101 & 0101 & 1101 & 1100 & 1 \\
  1110 & 0110 & 1011 & 1010 & 1 \\
  1111 & 0111 & 0000 & 0001 & 0 \\
  \hline
\end{tabular}
\end{center}
\caption{A linear relation for the key bits ($b$ arises from $a$ by adding
   $k^{(0)}$, resulting in ``flipping'' the first bit, $b'$ from $b$ by applying
   $f$, and $c$ from $b'$ by adding $k^{(1)}$.}\label{tab-bool-linrel}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|cccccccccccccccc|} \hline
     & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 &10 &11 &12 &13 &14 &15 \\ \hline
   0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
   1 & 0 &$-\frac{1}{4}$&$-\frac{1}{4}$& 0 & 0 &$-\frac{1}{4}$&$-\frac{1}{4}$& 0 & 0 &$-\frac{1}{4}$&$-\frac{1}{4}$& 0 & 0 &$\frac{3}{4}$&$-\frac{1}{4}$& 0 \\
   2 & 0 &$\frac{1}{4}$& 0 & $-\frac{1}{4}$ &$-\frac{1}{2}$& $-\frac{1}{4}$ & 0 & $-\frac{1}{4}$ & $-\frac{1}{4}$ &$\frac{1}{2}$& $-\frac{1}{4}$ & 0 &$\frac{1}{4}$& 0 & $-\frac{1}{4}$ & 0 \\
   3 & 0 &$\frac{1}{2}$&$\frac{1}{4}$& $-\frac{1}{4}$ &$\frac{1}{2}$& 0 &$\frac{1}{4}$& $-\frac{1}{4}$ & $-\frac{1}{4}$ & $-\frac{1}{4}$ & 0 & 0 &$\frac{1}{4}$&$\frac{1}{4}$& 0 & 0 \\
   4 & 0 & 0 &$-\frac{1}{2}$& 0 & 0 & 0 & 0 &$-\frac{1}{2}$&$\frac{1}{4}$& $-\frac{1}{4}$ & $-\frac{1}{4}$ & $-\frac{1}{4}$ &$\frac{1}{4}$& $-\frac{1}{4}$ &$\frac{1}{4}$&$\frac{1}{4}$\\
   5 & 0 &$\frac{1}{4}$&$\frac{1}{4}$&$\frac{1}{2}$& 0 &$\frac{1}{4}$& $-\frac{1}{4}$ & 0 &$\frac{1}{4}$& 0 &$-\frac{1}{2}$&$\frac{1}{4}$&$\frac{1}{4}$& 0 & 0 & $-\frac{1}{4}$ \\
   6 & 0 &$\frac{1}{4}$& 0 &$\frac{1}{4}$& 0 &$\frac{1}{4}$& 0 &$\frac{1}{4}$& 0 &$\frac{1}{4}$& 0 &$-\frac{3}{4}$& 0 &$\frac{1}{4}$& 0 &$\frac{1}{4}$\\
   7 & 0 & 0 &$\frac{1}{4}$& $-\frac{1}{4}$ & 0 & 0 &$-\frac{3}{4}$& $-\frac{1}{4}$ & 0 & 0 &$\frac{1}{4}$& $-\frac{1}{4}$ & 0 & 0 &$\frac{1}{4}$& $-\frac{1}{4}$ \\
   8 & 0 & 0 & $-\frac{1}{4}$ &$\frac{1}{4}$& $-\frac{1}{4}$ &$\frac{1}{4}$& 0 & 0 &$-\frac{1}{2}$& 0 &$\frac{1}{4}$&$\frac{1}{4}$&$\frac{1}{4}$&$\frac{1}{4}$&$\frac{1}{2}$& 0 \\
   9 & 0 &$\frac{1}{4}$& 0 &$\frac{1}{4}$& $-\frac{1}{4}$ &$-\frac{1}{2}$&$\frac{1}{4}$& 0 & 0 & $-\frac{1}{4}$ & 0 & $-\frac{1}{4}$ & $-\frac{1}{4}$ & 0 &$\frac{1}{4}$&$-\frac{1}{2}$\\
  10 & 0 & $-\frac{1}{4}$ &$\frac{1}{4}$& 0 & $-\frac{1}{4}$ & 0 & 0 &$\frac{1}{4}$& $-\frac{1}{4}$ &$-\frac{1}{2}$& 0 & $-\frac{1}{4}$ &$\frac{1}{2}$& $-\frac{1}{4}$ & $-\frac{1}{4}$ & 0 \\
  11 & 0 &$\frac{1}{2}$& 0 & 0 & $-\frac{1}{4}$ & $-\frac{1}{4}$ & $-\frac{1}{4}$ &$\frac{1}{4}$&$\frac{1}{4}$& $-\frac{1}{4}$ &$\frac{1}{4}$&$\frac{1}{4}$& 0 & 0 & 0 &$\frac{1}{2}$\\
  12 & 0 & 0 &$\frac{1}{4}$&$\frac{1}{4}$& $-\frac{1}{4}$ &$\frac{1}{4}$& 0 &$-\frac{1}{2}$& $-\frac{1}{4}$ & $-\frac{1}{4}$ & 0 & 0 &$-\frac{1}{2}$& 0 & $-\frac{1}{4}$ &$\frac{1}{4}$\\
  13 & 0 & $-\frac{1}{4}$ &$\frac{1}{2}$& $-\frac{1}{4}$ & $-\frac{1}{4}$ & 0 &$\frac{1}{4}$& 0 &$\frac{1}{4}$& 0 & $-\frac{1}{4}$ & 0 & 0 &$\frac{1}{4}$&$\frac{1}{2}$& 0 \\
  14 & 0 & $-\frac{1}{4}$ &$\frac{1}{4}$&$\frac{1}{2}$&$\frac{1}{4}$&$-\frac{1}{2}$& 0 & $-\frac{1}{4}$ & 0 &$\frac{1}{4}$&$\frac{1}{4}$& 0 &$\frac{1}{4}$& 0 & 0 &$\frac{1}{4}$\\
  15 & 0 & 0 & 0 & 0 &$\frac{1}{4}$& $-\frac{1}{4}$ & $-\frac{1}{4}$ &$\frac{1}{4}$&$-\frac{1}{2}$& 0 &$-\frac{1}{2}$& 0 & $-\frac{1}{4}$ & $-\frac{1}{4}$ &$\frac{1}{4}$&$\frac{1}{4}$\\ \hline
\end{tabular}
\end{center}\caption{Correlation matrix of the S-box $\mathrm{S}_0$ of {\sc Lucifer}\index{Lucifer}.
   Row and column indices are linear forms represented by integers.}\label{tab-bool-corr}
\end{table}

@@

     & 0            &$\frac{1}{16}$& 0            & 0            &$\frac{1}{16}$&$\frac{1}{4}$ & 0 \\

@@

Once again we consider a concrete example with $m = 2$ and $q = 4$,
hence $n = 8$. As S-boxes we take the ones from {\sc Lucifer}, $\textrm{S}_0$
at the left, and $\textrm{S}_1$ at the right, see Figure~\ref{fig-bool-par}.
For the left S-box $\textrm{S}_0$ we take the linear relation
with $\alpha \:\hat{=}$ \verb:0001: and $\beta \:\hat{=}$ \verb:1101:,
that we know has probability $p_1 = \frac{7}{8}$. For the right
S-Box $\textrm{S}_1$ we take the relation $(0,0)$ (where both linear forms are $0$).
Since always $0 = 0$ its probability is $1$.
The combined linear relation for $f = (\textrm{S}_0,\textrm{S}_1)$ then
also has probability $p = \frac{7}{8}$ and potential $\lambda = \frac{9}{16}$
by Theorem~\ref{thm-bool-parallel},
and we know that linear cryptanalysis with $N = 5$ pairs of plaintext
and ciphertext has (more than) 95\% success probability. We decompose all relevant
bitblocks into bits:
\begin{description}
   \item[plaintext:] $a = (a_0, \ldots, a_7) \in \F_2^8$,
   \item[ciphertext:] $c = (c_0, \ldots, c_7) \in \F_2^8$,
   \item[key:] $k = (k_0, \ldots, k_{15}) \in \F_2^{16}$ where
      $(k_0, \ldots, k_7)$ serves as ``initial key'' (corresponding to
      $k^{(0)}$ in Figure~\ref{fig-bool-par}), and $(k_8, \ldots, k_{15})$
      as ``final key'' (corresponding to $k^{(1)}$).
\end{description}
Then $\alpha(a) = a_3$, $\beta(c) = c_0 + c_1 + c_3$, and
$\kappa(k) = \alpha(k_0, \ldots, k_7) + \beta(k_8, \ldots, k_{15}) = k_3 + k_8 + k_9 + k_{11}$.
Hence the target relation is
\[
     k_3 + k_8 + k_9 + k_{11} = a_3 + c_0 + c_1 + c_3.
\]
We use the key $k =$ \verb:1001011000101110: whose relevant bit is
$k_3 + k_8 + k_9 + k_{11} = 1$, and generate five random pairs of
plaintext and ciphertext\footnote{%
   We simulate an attacker who knows five pairs of plaintexts and corresponding ciphertexts
   by generating (arbitrarily at random) and using these five pairs.
   Our odds of correctly determining the wanted key bit are very high.
}, see Table~\ref{tab-bool-BspC}. We see that
for this example Matsui's algorithm guesses the relevant key bit
correctly with no dissentient.

@@

\begin{figure}[h] % [htbp]
\begin{center}
\begin{picture}(320,50)
  \linethickness{2pt}
  \put(20,20){\line(1,0){260}}
  \put(20,20){\line(0,1){30}}
  \put(20,50){\line(1,0){260}}
  \put(280,20){\line(0,1){30}}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\section{Appendix: Boolean Maps in SageMath}\label{a-bool-sage}
\subsection{What's in SageMath?}
